Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-17 Direction.tex
Text from Matt and update the acknowledgements. --Lindley

%%idea: use (some) PMTs with high QE for high lambda (and/or low QE for normal lambda). 
%% add some more info why we are interested in a smaller detector
%% motivation: Daedalus/ISODAR (?), 0nuBB 
\section{Reconstruction}
\label{reconstruction_sec}
Hopefully results

@@

\section{Acknowledgements}
The authors would like to thank H. Frisch for useful discussion and comments in preparing this work. The authors would like to thank K. Arisaka for discussion on the capabilities of traditional PMTs in addition to HPDs. Finally, L. Winslow would like to thank J.M. Conrad for suggesting that the LAPPD may have sufficient timing resolution to allow the reconstruction of the electron direction in $0\nu\beta\beta$ and that a collaboration with the University of Chicago should be pursued.  A. Elagin and M. Wetstein are supported by DOE XXXX. C. Aberle and L. Winslow are supported by funds from University of California Los Angeles.

>>>>

%%idea: use (some) PMTs with high QE for high lambda (and/or low QE for normal lambda). 
%% add some more info why we are interested in a smaller detector
%% motivation: Daedalus/ISODAR (?), 0nuBB 
\section{Reconstruction}
\label{reconstruction_sec}
WCSimAnalysis is a Water Cherenkov reconstruction package developed by Andrew Blake for the Long Baseline Neutrino Experiment (LBNE collaboration). It provides a framework for generic event cleaning, track reconstruction, and particle id, and it comes equipped with variety of pre-built algorithms. A collaboration between Iowa State, the University of Chicago, and Argonne National Laboratory has continued adding to the code, developing new track-fitting techniques for Water Cherenkov detectors built around the unique capabilities of microchannel plate photodetectors (MCPs): timing resolutions below 100 picoseconds and imaging resolutions below a centimeter. In this paper, we use only the simple, low-level reconstruction tools native to the package, though in future works we hope to implement more sophisticated techniques.

The simplest approach to vertex reconstruction, used in this paper, is commonly referred to as a “point fit”. It assumes that, to first order, the light was emitted a single point in space-time (x0,y0,z0,t0). In reality, the light is emitted along the extended, multi-scattered track of the electron. However, at the energies discussed in this paper, the extent of the electron track is small (a few cm) compared to the scale of the detector, which has a radius of 6 meters. It is also small compared with the vertex resolutions typical of experiments with this scale. As starting point, the assumption works well.

The first algorithm we use to find a seed vertex relies on exact numerical calculations of possible vertex candidates. Given a single point source, we need four constraints to solve for the four unknowns of the vertex. A collection of N random quadruplets of hits is selected and used to solve for N corresponding vertex candidates. Some of these solutions will produce nonsensical results due to optical scattering of the emitted light, slow decay constant of the scintillation, finite detector resolution, and the limitations of our assumption of a point source. Nonetheless, over the N unique quadruplets, we expect some of the solutions to lie near the true event vertex. In this paper we chose N to be 80, determined by trial and error. However, given many hundreds of hits, the number of possible 4-hit combinations is much larger and one can choose a different N to better suit one’s particular experimental context.

Once the 80 vertex candidates have been solved for, we test the goodness of each candidate and select the one that best fits the full ensemble of hits. Distances are calculated between each hit and the candidate point-source. From these distances, we calculate the transit time, given the speed of light in our scintillator medium. This transit time plus the t0 of the 4-vertex provides a predicted arrival time for each hit. The “time residual” is the calculated as difference between the predicted and the measured arrival time, given the vertex candidate. Since we assume a point-source in this paper, we use what we call the “simple time residual”, as distinct from the “extended time residual”, which assumes an extended, track-like geometry. The distribution of time residuals for each hit with respect to a vertex candidate has a width that is minimized when the hypothesis vertex is close to the true vertex. The t0 of the vertex candidate is left to float, and is adjusted to maximize the likelihood comparison with a Gaussian test function. The width of that gaussian test function is hand optimized to correspond with the width of the “best fit”. The vertex candidate with the largest maximum is selected to serve as the best candidate seed vertex. Once a single best seed vertex has been chosen, the same time-residual figure-of-merit is used in a MINUIT-based fit over a continuum of possible vertices in the neighborhood of the seed.

Once the vertex fitting is completed, we determine the direction by taking the centroid of all vectors pointing from the fitted event vertex and the hits on the detector. Since our initial time cut greatly enhances the purity of the Cherenkov light in this sample of hits and, since the Cherenkov light is directional (compared to the isotropic scintillation light), this fit provides a good measure of the track direction. This approach does not rely on the highly constrained Cherenkov geometry of the event, which contains further information. Future work will focus on better algorithms. Nonetheless, for a proof-of-principles, this method works sufficiently well..

@@

\section{Acknowledgements}
L. Winslow would like to thank for useful discussion on the topic. L. Winslow would also like to thank K. Arisaka for discussions on the possible reach of traditional PMTs and the characteristics of HPDs. A. Elagin, H.J. Frisch and M. Wetstein are supported by DOE XXXX. C. Aberle and L. Winslow are supported by funds from University of California Los Angeles.
